Characterizing subarctic biomes for land surface modeling of pollution and climate risk.

### 2021-02-08 VegOzone.tex
First suggested changes by Lisa. List with todos for introduction.

\begin{abstract}
Here we assess the importance of characterization of key meteorological variables and air pollution profiles on the modeled susceptibility of subarctic vegetation to ozone. Changes in climate (e.g. temperature, water availability) will manifest in both direct and indirect effects influencing the native species sensitivity to air pollution. A likely increase in frequency of persistent heatwaves in the future, will inflict heat related stress while generating favorable conditions for the formation of ambient air ozone, e.g. due to wildfires.
We have monitored ambient air ozone concentrations at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd in Northern Norway in the vegetation period 2018/19. Due to data acquisition problems, ozone concentrations for two weeks in July 2018 were missing from our record. We present a reconstruction based on probability density function with respect to the Swedish and Finnish atmospheric monitoring sites and compare with available surface ozone products for the region. Over all, ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance diverged substantially from the multi annual mean precipitation varied only to some extend. 
We investigate the difference in uptake of ozone using the DO3SE model, for typical vegetation at the location. We find that bespoke parameterization for subarctic vegetation is essential for an accurate estimation of Phototoxic Ozone Dose (POD) and an assessment of damage risk. Accurate estimates of the start and length of growing season are also crucial.
We find that temperature and its derivative vapor pressure deficit are likely the most important climate variables while the variability of ozone concentration within climatologically reasonable limits plays a minor role.
The use of subarctic parameterization suggests an improvement in risk assessment as POD values are increased in 2018; corroborated by evidence of visible injury in the Svanhovd ozone garden.
\end{abstract}

@@

Tropospheric and in particular ambient air ozone is a highly toxic pollutant to all ecosystems. In continuously high concentration, ozone negatively affects life expectancy of humans \citep{WHO2008} and is cause to early senescence as well as programmed cell death in plants \citep{PCE:Kangasjarvi2005}. On average, a global loss of yield in four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. Former can proof a vital threat to food security in fast developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone related to air pollution has increased in the past years. At the same time, a successful implementation of air quality regulations reduced the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}.

With a radiative forcing of $0.40 \pm 0.20\,\unit{W\,m^{-2}}$, ozone is a potent climate forcer, ranked third behind \chem{CO_2} and \chem{CH_4} \citep[Chapter 8]{IPCC2013}. Its damaging effects on global vegetation might provide an important additional feedback to climate change \citep{Nat:Sitch2007}.

In the presents of nitrogen oxides (\chem{NO_x}), tropospheric ozone is formed in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic substances (VOCs). In the absence of local pollution sources, episodes of high ozone concentrations are often attributed to long-range transport of precursors. In northern Fennoscandia, episodes in 2003 and 2006 have been traced back to emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Hydrocarbons are also emitted by living vegetation in form of terpenes and monoterpenes, so called biogenic volatile substances (BVOCs). Those emissions are associated with, e.g., thermal stress, defense against herbivores, or attraction of pollinators \citep{TPS:Penuelas2003}.

Despite a relatively short tropospheric life-time of roughly $22\,\unit{days}$, ranging from a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}, ozone is subject to advection over long distances.
In the annual cycle, a distinct maximum typically occurs in spring followed by lower concentrations throughout the summer. This is most likely due to stronger dry deposition on land vegetation than on bare ground or snow and ice \citep{AE:Monks2000}. This spring peak is reached earlier in northern Fennoscandia than in more southern locations but is highly variable throughout the years \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is a complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

At the same time, the start of growing season in northern Fennoscandia is shifting to earlier days of the year in response to the general warming trend \citep[e.g.,]{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007} and thereby into close vicinity of the ozone spring peak. This may promote a higher susceptibility of deciduous species to ozone.

Since the industrial revolution, tropospheric background concentrations of ozone in the northern hemisphere have doubled and kept increasing until recently. But long-term observations started before 1950 are scarce \citep[Chapter 2]{IPCC2013}. Recent trend analyses indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level-off or even decline \citep{ACP:Wespes2018,ESA:Gaudel2018}.\\

\emph{Contributor: Ane} {\bf TODO: More detailed ozone damage description?}\\
Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.

The opening of stomata, quantified through their conductance to water vapor $g_\mathrm{sto}$, is governed by different environmental factors and can be described by a multiplicative ansatz \citep{PTRS:Jarvis1976,ICP:MappingManual2017}:
\begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within a range $0-1$. They are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type (PFT) $k$.

>>>>

\begin{abstract}
Here we assess the importance of characterization of key meteorological variables and air pollution profiles on the modeled susceptibility of subarctic vegetation to ozone. Changes in climate (e.g. temperature, water availability) will manifest in both direct and indirect effects influencing the species' sensitivity to air pollution. A likely increase in frequency of persistent heatwaves in the future, will inflict heat related stress while generating favorable conditions for the formation of ambient air ozone, e.g. due to wildfires.
We have monitored ambient air ozone concentrations at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd in Northern Norway in the vegetation period 2018/19. Due to data acquisition problems, ozone concentrations for two weeks in July 2018 were missing from our record. We present a reconstruction based on probability density function with respect to the Swedish and Finnish atmospheric monitoring sites and compare with available surface ozone products for the region. Over all, ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance diverged substantially from the multi annual mean precipitation varied only to some extend. 
We investigate the difference in uptake of ozone using the DO3SE model, for typical vegetation at the location. We find that bespoke parameterization for subarctic vegetation is essential for an accurate estimation of Phototoxic Ozone Dose (POD) and an assessment of damage risk. Accurate estimates of the start and length of growing season are also crucial.
We find that temperature and its derivative vapor pressure deficit are likely the most important climate variables while the variability of ozone concentration within climatologically reasonable limits plays a minor role.
The use of subarctic parameterization suggests an improvement in risk assessment as POD values are increased in 2018; corroborated by evidence of visible injury in the Svanhovd ozone garden.
\end{abstract}

@@

\begin{itemize}
\item Wealth of evidence demonstrating ozone damage to vegetation (perhaps focus on key evidence from northern climates and key species (e.g. birch, Norway srpuce etc... - this might help to make it clear why we focus on these species with the PODy modelling))
\item Although O3 trends stagnant or decreasing in Europe there is unique photochemcistry over the arctic (causing spring time peaks in ozone) that may mean ozone damage continues to be an issue
\item The earlier timing of growing seasons with climate change and 24 hour sunlight may also enhance the exposure of plants during those times when they are physiologically active to higher [O3]
\item Research also clearly shows how ecosystem sensitivity to vegetation is influenced by climate variables, both because this can influence ozone uptake (dose) as well as the impact of altered ozone doses (can add in some relevant detail here on ozone damage mechanisms - again perhaps with a focus on what we know for colder climates)
\item Risk assessment methods have been developed that allow for ozone uptake (i.e. PODy)
\item Comparison of these methods with previously used concentration based methods (e.g. AOT40) show that PODy is better able to represent the risk to vegetation from ozone, especially in the more climate extreme parts of Europe (see Mills et al 2011 paper attached - that also includes some work for clover that we could use)
\item These methods rely on accurate representation of plant physiological response to climate variables (i.e. the species or plant funtional type specific responses used to parameterise risk assessment models)
\item These models are most often parameterised using data for less extreme climates (i.e. temperate or continental climates; and to a lesser extent, boreal and Mediterranean climates. Parameterisation of these models for more extreme bio-geographical regions such as sub arctic climates has not been perfromed and relies on assuming that boreal parameterisations will be applicable.
\item It is important to understand the implications of using these more generic parameterisations for more extreme climate regions, here we explore how risk estimates might be over or underestimated using generic vs bespoke sub-arctic parameterisations
\item we also explore the role [O3] profile and changing climate variables to understand how this risk might alter in the future with the advent of climate change 
\item This allows both an understanding of how the threat from ozone may change over time but also which are the key aspects of plant physiology that might determine risk (and hence which parameters might require special attention to get right) both for pollution impact assessment, but also more generally for describing gas exchange response to changing climates that will influence bio-geochemcial cycling (could we add something in here about the paucity of PFTs representing subarctic conditions in biogeochemcial models?).
\end{itemize}

Ground level ozone is a highly toxic pollutant known to cause damage to a variety of ecosystems around the World \citep{PT:Emberson2020}. At continuously high concentration, ozone negatively affects life expectancy of humans \citep{WHO2008} and is a cause of visible injury, phtosynthetic damage, early senescence as well as programmed cell death \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone related to enhanced pollutant emissions from the transport, residential, power generation and industrial sectors have increased in the past few decades. At the same time, a successful implementation of air quality regulations coupled with economic restructuring which saw a move from coal to gas, reduced the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}.

With a radiative forcing value of $0.40 \pm 0.20\,\unit{W\,m^{-2}}$, ozone is a potent climate forcer, ranked third behind \chem{CO_2} and \chem{CH_4} \citep[Chapter 8]{IPCC2013}. Its damaging effects on the carbon sequestered by global vegetation might provide an important additional feedback to climate change \citep{Nat:Sitch2007}.

In the presence of nitrogen oxides (\chem{NO_x}), tropospheric ozone is formed in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic substances (VOCs). In the absence of local pollution sources, episodes of high ozone concentrations are often attributed to long-range transport of precursors. In northern Fennoscandia, ozone episodes in 2003 and 2006 have been traced back to \chem{NO_x} and \chem{VOC} emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Hydrocarbons are also emitted by living vegetation in form of terpenes and monoterpenes, so called biogenic volatile substances (BVOCs). These emissions are associated with, e.g., thermal stress, defense against herbivores, or attraction of pollinators \citep{TPS:Penuelas2003}.

Despite a relatively short tropospheric life-time of approximately $22\,\unit{days}$, ranging from a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}, ozone is subject to advection over long distances.
In the annual cycle in boreal and temperate climates, a distinct maximum typically occurs in spring followed by lower concentrations throughout the summer. This is most likely due to stronger dry deposition on land vegetation than on bare ground or snow and ice \citep{AE:Monks2000} causing a build off of ozone and its precursors in the course of polar night. Come spring they are photochemical activated again. This spring peak is reached earlier in northern Fennoscandia than in more southern locations but is highly variable throughout the years \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is a complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

At the same time, the start of growing season in northern Fennoscandia is shifting to earlier days of the year in response to the general warming trend \citep[e.g.,]{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007} and thereby converging with the period of the ozone spring peak. This may promote a higher susceptibility of deciduous species to ozone and could also affect coniferous species which will have flushes of new needles in the spring time which are likely, through their immaturity, to be more susceptible to ozone damage.

Even though long-term (pre 1950) observations of surface ozone concentrations are scarce; evidence from trend analysis of those few sites that have monitored concentrations suggests that since the industrial revolution, background concentrations in the northern Hemisphere have at least doubled and continued to increase until relatively recently \citep[Chapter 2]{IPCC2013}. Recent trend analyses indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level-off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}.\\

Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange allowing the entry of CO2 for photosynthesis and loss of H2O vapour via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the course of the day and growing season primarily to balance CO2 uptake against H2O vapour  loss.  The higher the stomatal conductance, the higher the potential for ozone uptake.\\

{\bf TODO: Move to Section~\ref{sec:do3se}.}
The opening of stomata, quantified through their conductance to water vapor $g_\mathrm{sto}$, is governed by different environmental factors and can be described by a multiplicative ansatz \citep{PTRS:Jarvis1976,ICP:MappingManual2017}:
\begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within a range $0-1$. They are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type (PFT) $k$.
